\documentclass[]{article}

\usepackage{fullpage}
\usepackage{physics}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{url}
\usepackage{comment}

%opening
\title{Implementation Notes for Gyrokinetic Particle Pusher}
\author{Norman M. Cao}

\begin{document}

\maketitle

\section{Equations of motion and geometry}

\subsection{Gyrokinetic characteristic equations (electrostatic)}
We aim to push particles in electrostatic gyrokinetics.
The equations of motion are given by
\begin{subequations}
	\begin{gather}
		B_\parallel^* \dot{\vb{R}} = \frac{1}{q} \vu{b} \times \nabla H + v_\parallel \vb{B}^* \\
		B_\parallel^* \dot{p}_\parallel = -\vb{B}^* \cdot \nabla H \\
		\dot{\mu} = 0
	\end{gather}
\end{subequations}
With some definitions
\begin{subequations}
	\begin{gather}
		\vu{b} := \vb{B} / B \\
		\vb{B}^* := \vb{B} + \nabla \times (p_\parallel \vu{b} / q) \\
		B_\parallel^* := \vu{b} \cdot \vb{B}
	\end{gather}
\end{subequations}
\begin{subequations}
	\begin{gather}
		H = p_\parallel^2 / 2 m + \mu B + q \mathcal{J}[\Phi] \\
		v_\parallel := \partial_{p_\parallel} H = p_\parallel / m
	\end{gather}
\end{subequations}
and \(m,q\) are the species mass and charge respectively.

\subsection{Gyrokinetic characteristic equations (electromagnetic)}
\begin{subequations}
	\begin{gather}
		B_\parallel^* \dot{\vb{R}} = \frac{1}{q_s} \vu{b} \times \pdv{H}{\vb{R}} + \frac{\vb{B}^*}{m_s} \pdv{H}{u_\parallel} \\
		B_\parallel^* \dot{u}_\parallel = - \frac{\vb{B}^*}{m_s} \cdot \pdv{H}{\vb{R}} - \frac{q_s B_\parallel^*}{m_s} \pdv{A^{(s)}_\parallel}{t}
	\end{gather}
\end{subequations}
\begin{subequations}
	\begin{gather}
		H = \frac{1}{2 m_s} (m_s u_\parallel - q_s A_\parallel^{(h)})^2 + \mu B + q_s \Phi \\
		\pdv{H}{\vb{R}} = \mu \nabla B + q_s \nabla \Phi - \frac{q_s}{m_s}(m_s u_\parallel - q_s A_\parallel^{(h)}) \nabla A_{\parallel}^{(h)} \\
		\pdv{H}{u_\parallel} = m_s u_\parallel - q_s A_{\parallel}^{(h)} \\
		\vb{B}^* = \vb{B} + \nabla \times ([m_s u_\parallel/ q_s + A_\parallel^{(s)}] \vu{b})
	\end{gather}
\end{subequations}

In the current implementation, we only include the Hamiltonian portion \(A_\parallel^{(h)}\) in the particle pusher.
Note that the generalized potential \(\Phi \approx \phi - \frac{m_s}{2 q_s} |\vb{v}_E|^2\) where \(\vb{v}_E = \frac{\vu{b} \times \nabla \phi}{B}\) is also written here for completeness, although we do not use it.

\subsection{Cylindrical coordinates}
We primarily work in a right-handed \((R,\varphi,Z)\) cylindrical coordinate system.
Thus, \(\varphi\) points into the \((R,Z)\) plane.
Recall that for a path \(\vb{R} = (R(t), \varphi(t), Z(t))\), we have that
\begin{gather*}
	\dot{\vb{R}} = \dot{R} \vu{R} + \dot{Z} \vu{Z} + R \dot{\varphi} \vu*{\varphi} \\
	\ddot{\vb{R}} = (\ddot{R} - R \dot{\varphi}^2) \vu{R} + \ddot{Z} \vu{Z} + (R \ddot{\varphi} + 2 \dot{R} \dot{\varphi}) \vu*{\varphi}
\end{gather*}
Typically we will work with the orthonormal basis of unit vectors \((\vu{R}, \vu*{\varphi}, \vu{Z})\).
From the above expressions, the velocity acts on coordinates as
\begin{equation*}
	\dot{R} = \dot{\vb{R}} \cdot \vu{R} \qquad
	\dot{Z} = \dot{\vb{R}} \cdot \vu{Z} \qquad
	\dot{\varphi} = (\dot{\vb{R}} \cdot \vu*{\varphi}) / R
\end{equation*}
then, the acceleration acts on components of the velocity vector (in the orthonormal basis) as
\begin{gather*}
	 \\
	(\dot{\vb{R}} \cdot \vu{R})' = \ddot{\vb{R}}\cdot \vu{R} + R \dot{\varphi}^2 \\
	(\dot{\vb{R}} \cdot \vu{Z})' = \ddot{\vb{R}} \cdot \vu{Z} \\
	(\dot{\vb{R}} \cdot \vu*{\varphi})' = \dot{R} \dot{\varphi} + R \ddot{\varphi} = \ddot{\vb{R}} \cdot \vu*{\varphi} - \dot{R}\dot{\varphi}
\end{gather*}

\subsection{Magnetic field}
We use the following representation of the magnetic field and current
\begin{gather*}
	\vb{B} = F(\psi) \nabla \varphi + \nabla \varphi \times \nabla \psi = \frac{F(\psi) \vu*{\varphi} + \vu*{\varphi} \times \nabla \psi}{R} \\
	\nabla \times \vb{B} = F' \nabla \psi \times \nabla \varphi + \nabla \times (\nabla \varphi \times \nabla \psi)
\end{gather*}
where \(\psi\) is the poloidal flux.
Note the second term in the current can be evaluated using an in-plane curl

It's useful to have the following representations for certain terms in the gyrokinetic equation:
\begin{gather*}
	\nabla B = \frac{\nabla(RB) - B \nabla R}{R} \\
	2 RB \nabla(RB) = \nabla(R^2B^2) = \nabla(F^2 + |\nabla\psi|^2) = 2 F' \nabla \psi + 2 \operatorname{Hess}[\psi] \nabla \psi \\
	\nabla \times \vu{b} = \frac{B (\nabla \times \vb{B}) - (\nabla B) \times \vb{B}}{B^2}
\end{gather*}
note that these can be written purely in terms of analytic derivatives of \(\psi(R,Z)\) and \(F(\psi)\).

\section{Eikonal Representation of Fields}

\subsection{Flux Coordinates}
First, we define the flux coordinates that we adopt in order to represent modes.
We use \(\psi\) as the radial coordinate and \(\varphi\) as the toroidal coordinate.
Now, we define a binormal coordinate \(\theta\) by a ``UV mapping'' strategy.
Namely, we define fields \((u(R,Z), v(R,Z))\) such that \(\theta = \arctan(v/u)\).
These UV fields are defined to satisfy the following:
\begin{enumerate}
	\item \(u=v=0\) at the magnetic axis
	\item \(\nabla \theta = 0\) at the two primary magnetic X-points
	\item At each magnetic null, the Hessian \(H[\psi]\) is congruent to a diagonal matrix \(D\) with entries \(\pm 1\), \(H[\psi] = S D S^T\); this is Sylvester's law of inertia. We constrain the Hessian \(H[\theta] = S A S^T\) where \(A_{11} = A_{22} = 0\) and \(A_{12}=A_{21} = a\).
\end{enumerate}
These help ensure that \(\theta\) acts as a good ``conjugate coordinate'' to \(\psi\).

\subsection{Clebsch Representation of Magnetic Field}

Consider a Clebsch representation of the magnetic field
\begin{equation*}
	\vb{B} = \nabla \psi \times \nabla \alpha
\end{equation*}
Any function \(S(\psi,\alpha)\) automatically satisfies \(\vb{B} \cdot \nabla S = 0\).
We can choose \(\alpha = \alpha_\Phi(\psi,\theta) - \varphi\).
Note that \(\alpha_\Phi\) is the component related to the toroidal flux.

To compute \(\alpha_\Phi\), notice that
\begin{equation*}
	\vb{B} = \pdv{\alpha_\Phi}{\theta} \nabla \psi \times \nabla \theta + \nabla \varphi \times \nabla \psi
\end{equation*}
In other words,
\begin{equation*}
	\pdv{\alpha_\Phi}{\theta} = \frac{\vb{B} \cdot \nabla \varphi}{(\nabla \psi \times \nabla \theta) \cdot \nabla \varphi}
\end{equation*}
Thus, by integrating along curves of constant \(\psi\) and varying \(\theta\), it is possible to compute \(\alpha_\Phi\).

We make a brief remark about singularities.
Let \(\vb{x} = (R-R_0, Z-Z_0)\), and let \(\vb*{\xi} = S^T \vb{x}\).
Near poloidal magnetic field nulls, we have
\begin{gather*}
	\psi \approx \frac{1}{2} \vb{x}^T H[\psi] \vb{x} = \frac{1}{2} \vb*{\xi}^T D \vb*{\xi} \\
	\theta \approx \frac{1}{2} \vb{x}^T H[\theta] \vb{x} = \frac{1}{2} \vb*{\xi}^T A \vb*{\xi}
\end{gather*}

\subsection{Eikonal Ansatz}

We take an eikonal ansatz of the form
\begin{equation}
	\delta \phi = \exp(i(\omega t - n \alpha)) g(\alpha_\Phi, \psi)
\end{equation}
In the JAX version, automatic differentiation is used to compute gradients of \(\delta \phi\).

\section{Monte Carlo Solution of Transport Equations}

In this section, we discuss the strategy for using Monte Carlo in combination with the Feynman-Kac formula to solve certain transport equations.

\subsection{Problem Setup}

First, we discuss our strategy for dealing with the collision operator.
Let's start with the nonlinear Landau collision operator, which is a Fokker-Planck operator of the form
\begin{equation}
	C_{ab} = \pdv{\vb{v}_a} \cdot \left[\vb{B}_{ab} f_a - \vb{D}_{ab} \cdot \pdv{f_a}{\vb{v}_a}\right]
\end{equation}
See \url{https://farside.ph.utexas.edu/teaching/plasma/Plasma/node38.html} for definitions of \(\vb{B}_{ab}, \vb{D}_{ab}\) (note we have absorbed the mass factor into these definitions).

The gyrokinetic Vlasov equation can be written in conservative form
\begin{equation}
	\pdv{(B_\parallel^* f_a)}{t} + \nabla \cdot (\dot{\vb{R}} B_\parallel^* f_a) + \pdv{(\dot{u}_\parallel B_\parallel^* f_a)}{u_\parallel} = \sum_{b} B_\parallel^* C_{ab}^{gk}[f_a, f_b]
\end{equation}
where \(C_{ab}^{gk}\) is the gyroaveraged collision operator.
We assume that we can write this in Fokker-Planck form.

The key strategy we take is to ignore the conservation properties of the collision operator, and simply treat it as a velocity-space diffusion operator.
Similar to how the fields are not computed self-consistently from the particles, the collisional drag and diffusion are not computed self-consistently from the particle distributions.
Energy/momentum is not conserved in the test particle picture even in the collisionless case, as they exchange energy with the fields.
However, if the time-dependent fields were known, then the test particle picture would be exact.

Let's generalize this situation a bit.
Suppose \(F_s(z)\) satisfies the Fokker-Planck equation
\begin{equation*}
	\pdv{F_s}{t} + \pdv{\vb{z}} \cdot \left[\vb{b}_s[F] F_s - \vb{D}_s[F] \cdot \pdv{F_s}{\vb{z}}\right] = 0
\end{equation*}
where \(\vb{z}\) is the phase space coordinate, and \(\vb{b}_s, \vb{D}_s\) are the drift and diffusion coefficients respectively, depending on \(F = F_1, ..., F_n\).
Supposing that this nonlinear Fokker-Planck equation satisfies some (possibly nonlinear) conservation law \(\dv{t}Q[F] = 0\), we can consider
\begin{equation*}
	\pdv{G_s}{t} + \pdv{\vb{z}} \cdot \left[\vb{b}_s[H] G_s - \vb{D}_s[H] \cdot \pdv{G_s}{\vb{z}}\right] = 0
\end{equation*}
While in general we no longer expect \(\dv{t} Q[G] = 0\), the conservation law will still hold if \(H=G\).
Thus, we can think about solving just the transport part of the equation with a fixed background \(H\), and potentially iterating to self-consistency.

\subsection{Periodic Solutions to Transport Equations}
We now focus on
\begin{equation}
	\pdv{G}{t} + \pdv{\vb{z}} \cdot \left[\vb{b}(\vb{z}, t) G - \vb{D}(\vb{z}, t) \cdot \pdv{G}{\vb{z}}\right] = \pdv{G}{t} - \mathcal{L}(t)[G] = S(\vb{z}, t)
\end{equation}
subject to the boundary conditions
\begin{equation*}
	G = g(\vb{z}, t) \text{ on } \partial \Omega
\end{equation*}
where \(\vb{b}(\vb{z}, t), \vb{D}(\vb{z}, t), S(\vb{z}, t), g(\vb{z}, t)\) are periodic in time with period \(T\).
We want to find a periodic solution \(G(\vb{z}, t+T) = G(\vb{z}, t)\).

The first trick we do is to convert the inhomogeneous boundary conditions to homogeneous ones.
Let \(G = U + V\) where \(V\) is a periodic function which satisfies the boundary conditions, but not necessarily the PDE.
Then, \(U\) satisfies the PDE, but with homogeneous boundary conditions and a modified (but still time-periodic) source term, which we also call \(S\).
Thus, we focus on solving the homogeneous Dirichlet boundary condition case.

Let \(\mathcal{P}_{s}^{t}\) be the evolution operator taking the solution of the homogeneous Dirichlet problem from time \(s\) to time \(t\).
Starting with initial data \(U(0)\), we have (via Duhamel's principle)
\begin{equation*}
	U(t) = \mathcal{P}_{0}^{t}[U(0)] + \int_{0}^{t} \mathcal{P}_{\tau}^{t}[S(\tau)] \dd{\tau}
\end{equation*}
To enforce periodicity, we require \(U(0) = U(T)\), so we can write
\begin{equation}
	(\mathcal{I} - \mathcal{P}_{0}^{T}) U(0) = \int_{0}^{T} \mathcal{P}_{\tau}^{T}[S(\tau)] \dd{\tau}
\end{equation}
This equation allows us to solve for the initial data \(U(0)\) which result in a periodic solution.

Note that formally, \((\mathcal{I} - A)^{-1} = \sum_{m=0}^{\infty} A^m\) for any operator \(A\) with spectral radius less than 1.
This gives an explicit formal series representation of the solution:
\begin{align*}
	U(t) &= \mathcal{P}_{0}^{t} \sum_{m=0}^{\infty} (\mathcal{P}_{0}^{T})^m \left[\int_{0}^{T} \mathcal{P}_{\tau}^{T}[S(\tau)] \dd{\tau}\right] \\
	&= \int_{0}^{T} \sum_{m=0}^{\infty} \mathcal{P}_{\tau}^{t + (m+1)T} [S(\tau)] \dd{\tau} \\
\end{align*} 

Now, for kinetic problems, we are interested in computing temporally-averaged spatially-weighted velocity moments of \(G\); these will take the form of linear functionals of \(G\) that look like
\begin{equation*}
	\langle \eta, G \rangle = \int_{0}^{T} \int_{\Omega} \eta(\vb{z},t) G(\vb{z},t) \dd{\vb{z}} \dd{t}
\end{equation*}

Consider the adjoint to the Fokker-Planck equation, i.e. the Kolmogorov backward equation:
\begin{equation}
	-\pdv{w}{t} -\vb{b}(\vb{z},t) \cdot \pdv{w}{\vb{z}} - \pdv{\vb{z}} \cdot \left[\vb{D}(\vb{z},t) \cdot \pdv{w}{\vb{z}}\right] = \eta(\vb{z},t)
\end{equation}
with spatial boundary conditions \(w = 0\) on \(\partial \Omega\), and periodic in time \(w(0) = w(T)\).
Observe that
\begin{align*}
	\langle \eta, G \rangle
	&= \int_{0}^{T} \int_{\Omega} \left[-\pdv{w}{t} -\vb{b} \cdot \pdv{w}{\vb{z}} - \pdv{\vb{z}} \cdot \left(\vb{D} \cdot \pdv{w}{\vb{z}}\right)\right] G \dd{\vb{z}} \dd{t} \\
	&= \int_{0}^{T} \int_{\Omega} w \left[\pdv{G}{t} + \pdv{\vb{z}} \cdot \left(\vb{b} G - \vb{D} \cdot \pdv{G}{\vb{z}}\right)\right] \dd{\vb{z}} \dd{t} \\
	&= \int_{0}^{T} \int_{\Omega} w S \dd{\vb{z}} \dd{t} = \langle w, S \rangle
\end{align*}
In other words, solving the adjoint equation allows us to compute the desired moments of \(G\) without explicitly computing \(G\) itself.

\subsection{Monte Carlo Estimation}

The solution to the Kolmogorov backward equation can be represented as an expectation over stochastic trajectories.
For \(t > s\), let \(\vb{Z}_t(\vb{z},s)\) satisfy the foward-time SDE
\begin{equation}
	\begin{gathered}
		\dd{\vb{Z}_t} = \tilde{\vb{b}}(\vb{Z}_t, t) \dd{t} + \sigma(\vb{Z}_t, t) \cdot \dd{\vb{W}_t} \\
		\vb{Z}_s(\vb{z}, s) = \vb{z}
	\end{gathered}
\end{equation}
Then, we have the Feynman-Kac representation (TODO: check sign)
\begin{equation*}
	w(\vb{z}, s) = \mathbb{E}[w(\vb{Z}_T(\vb{z},s), T)] + \mathbb{E}\left[\int_{s}^{T} \eta(\vb{Z}_t(\vb{z}, s), t) \dd{t}\right]
\end{equation*}

\subsection{Computing Transport via Periodic Orbits}

Consider an observable that looks like \(\int_\Omega F(\vb{z}, t) \eta(\vb{z}) \dd{z}\)

Average over a finite time interval \([t_0, t_1]\) of IVP -> approximated by the `long-time' average <-> approximated by average over a low-diml chaotic attractor -> approximated by weighted average over smaller set of periodic orbits of the nonlinear GK-VP -> approximate nonlinear UPOs of GK-VP via periodic solutions to the passive transport problem (9), where \(\vb{b}(\vb{z},t)\) and \(\vb{D}(\vb{z},t)\) are computed from equilibrium + eigenmode perturbation.

\begin{equation*}
	\ev{A} = \sum_{\text{periodic orbits } p} w_p \ev{A}_p \approx \sum_{\text{some smaller set of periodic orbits } p} \tilde{w}_p \ev{A}_p
\end{equation*}


\section{\(C^1\) interpolation on unstructured meshes}

\subsection{Field line tracing}
Let \(\vec{R}_B(R,Z;\varphi) = (R_B(...), Z_B(...))\) be the motion of the field-line trace starting at \((R,Z)\) on a poloidal plane, parameterized by the toroidal angle \(\varphi\) moved along the field line.

\(\vec{R}_B\) satisfies the ODE
\begin{equation*}
	\dv{\vec{R}_B}{\varphi} = \vec{b}_p \circ \vec{R}_B := \eval{\frac{R\vb{B}_p}{B_t}}_{\vec{R}_B} = \eval{\frac{R \vu*{\varphi} \times \nabla \psi}{F(\psi)}}_{\vec{R}_B}; \qquad
	\vec{R}_B(R,Z;0) = (R,Z)
\end{equation*}
This ODE is essentially a reparameterization of the magnetic field line ODEs with \(\varphi\) as time.
For \(\dv{\varphi}\) we think of \((R,Z)\) as being parameters.
These ODEs have an associated variational equation
\begin{equation*}
	\dv{[D\vec{R}_B]}{\varphi} = ([D\vec{b}_p] \circ \vec{R}_B) [D\vec{R}_B]; \qquad D\vec{R}_B(R,Z;0) = I_{2\times2}
\end{equation*}
here we think of \(D\) as the differential in \((R,Z)\) with \(\varphi\) as a parameter, that is:
\begin{equation*}
	D\vec{b}_p = \begin{bmatrix}
		\partial_R(\vec{b}_p \cdot \vu{R}) & \partial_Z(\vec{b}_p \cdot \vu{R}) \\
		\partial_R(\vec{b}_p \cdot \vu{Z}) & \partial_Z(\vec{b}_p \cdot \vu{Z})
	\end{bmatrix}
\end{equation*}
\begin{equation*}
	D\vec{R}_B = \begin{bmatrix}
		\partial_R R_B & \partial_Z R_B \\
		\partial_R Z_B & \partial_Z Z_B
	\end{bmatrix}
\end{equation*}

\subsection{Field-aligned interpolation}
Suppose we are trying to interpolate \(\phi(R,\varphi,Z)\) knowing its values on some equally spaced poloidal planes \(\phi_i(R,Z)\).
This can be accomplished by
\begin{equation*}
	\phi(R,\varphi,Z) = \sum_{i} p_i(\varphi) \phi_i(\vec{R}_B(R,Z;\varphi_i-\varphi))
\end{equation*}
here \(p_i\) are some piecewise polynomial basis functions.
We can compute its gradient by
\begin{equation*}
	\nabla \phi = \sum_{i} \left[p' (\phi_i \circ \vec{R}_B) \nabla \varphi + p_i \nabla (\phi_i \circ \vec{R}_B)\right]
\end{equation*}
Using the chain rule,
\begin{align*}
	\nabla (\phi_i \circ \vec{R}_B) &=
	\begin{bmatrix}
		\nabla R & \nabla Z
	\end{bmatrix}
	[D(\phi_i \circ \vec{R}_B)]^T
	-
	\nabla \varphi \left(\dv{(\phi_i \circ \vec{R}_B)}{\varphi}\right) \\
	&=
	\begin{bmatrix}
		\vu{R} & \vu{Z}
	\end{bmatrix}
	[D \vec{R}_B]^T
	[[\nabla \phi_i] \circ \vec{R}_B]
	-
	\frac{\vu*{\varphi}}{R}\left([[\nabla \phi_i]\circ \vec{R}_B] \cdot \dv{\vec{R}_B}{\varphi}\right)
\end{align*}
Note that it's possible to show that \(\vb{B} \cdot \nabla (\phi_i \circ \vec{R}_B) = 0\).

\subsection{Choice of basis functions}

Traditional cubic spline interpolation, which minimizes `bending' and \(C^2\) continuity, has polynomial coefficients which are computed by solving a linear system involving all of the data points as well as boundary conditions.
Instead we rely on polynomial splines which involve only the 4 points in the neighborhood of any \(\varphi\), and generally enforce \(C^1\) continuity at the nodes.

Two options are considered.
The first is cubic Hermite interpolation, with an array of polynomial coefficients
\begin{equation*}
	p = \begin{bmatrix}
		0 & -1/2 & 1 & -1/2 \\
		1 & 0 & -5/2 & 3/2 \\
		0 & 1/2 & 2 & -3/2 \\
		0 & 0 & -1/2 & 1/2
	\end{bmatrix}
\end{equation*}
where \(p_i(t) = \sum_{j=0}^{3} p_{ij} t^j\) (here the array entries are being 0-indexed).
This scheme exactly interpolates the nodes and also enforces \(C^1\) continuity at the nodes with a value of the derivative given by the centered difference of the adjacent two nodes.

The second is a quadratic smoothing spline,
\begin{equation*}
	p = \begin{bmatrix}
		1/4 & -1/2 & 1/4 \\
		1/2 & 0 & -1/4 \\
		1/4 & 1/2 & -1/4 \\
		0 & 0 & 1/4
	\end{bmatrix}
\end{equation*}
This scheme can be thought of as the anti-derivative of linear interpolation on the derivative, computed via centered difference, while enforcing \(C^1\) continuity at the nodes.
The quadratic dependence sacrifices exact interpolation at the nodes in exchange for a derivative with less oscillations.

Finally we remark that parallel noise seems to be non-negligible; in theoretical cases, a Lanczos filter is applied along the field line to smooth out these high-frequency parallel fluctuations.


\subsection{Interpolation on poloidal planes}

Interpolation on poloidal planes uses rHCT elements, which are \(C^1\) elements that minimize a `bending energy'.
The code is essentially a fork of the matplotlib \texttt{CubicTriInterpolator}\footnote{\url{https://matplotlib.org/stable/api/tri_api.html#matplotlib.tri.CubicTriInterpolator}} with a few optimizations.

\section{Extraction of Ballooning Coefficients}

\subsection{Straight Field-line Coordinates}
On the closed flux surfaces, let \(\theta_g\) be the geometric poloidal angle relative to the magnetic axis, with \(\theta_g = 0\) representing the outboard midplane.
Using \(F(\psi) = B_t / R\), we can compute the relationship between the straight field-line angle \(\theta\) in terms of \(\theta_g\) by
\begin{gather*}
	\theta = \frac{1}{q(\psi)} \int_{\theta_0(\psi)}^{\theta_g} \frac{F(\psi)}{B_p(\psi,\theta_g')} \ell'(\theta_g') \dd{\theta_g'} \\
	q(\psi) = \frac{1}{2\pi} \int_{0}^{2\pi} \frac{F(\psi)}{B_p(\psi,\theta_g')} \ell'(\theta_g') \dd{\theta_g'}
\end{gather*}
where \(\ell'(\theta_g)\) is the derivative of the arclength of the field line along the flux surface with respect to \(\theta_g\), and \(\theta_0(\psi)\) is the arbitrary offset on each flux surface where \(\theta=0\) lies.
An easy choice is \(\theta_0(\psi) = 0\), which results in \(\theta=0\) being the outboard midplane.
This relationship is then numerically inverted to get \(\theta\).

\subsection{Ballooning Transform}

Moving to flux coordinates \((\psi, \zeta, \theta)\) with \(\zeta=\varphi\), we can always write
\begin{equation*}
	\phi(R,\varphi,Z) = \sum_{n=-\infty}^{\infty} e^{i n \zeta} \phi_n(\psi,\theta)
\end{equation*}
We can move to the covering space \(\theta \mapsto \eta\), let \(\rho = nq\), and introduce the eikonal factor
\begin{gather*}
	\phi_n(\psi, \theta) = \sum_{\ell=-\infty}^{\infty} \hat{\phi}_n(\psi, \eta + 2 \pi \ell) \\
	\hat{\phi}_n(\psi, \eta) = e^{-i\rho\eta} f_n(\rho,\eta) \\
	f_n(\rho, \eta) = \int_{-\infty}^{\infty} \dd{\theta_k} e^{i\rho\theta_k} \tilde{f}_n(\theta_k, \theta)
\end{gather*}
the complete representation is then
\begin{equation*}
	\phi(R,\varphi,Z) = \sum_{n=-\infty}^{\infty} \sum_{\ell = -\infty}^{\infty} \int_{-\infty}^{\infty} \dd{\theta}_k e^{in(\zeta - q(\eta - \theta_k + 2 \pi \ell))} \tilde{f}_n(\theta_k, \theta + 2 \pi \ell)
\end{equation*}

Now, if we had an exact ballooning symmetry, we would have for a fixed parameter \(\theta_k\) (related to the radial wavenumber),
\begin{equation*}
	f_n(\rho,\eta) = e^{i\rho\theta_k} \tilde{F}_n(\eta)
\end{equation*}
This is exactly analogous to \(f(x) = \tilde{F} e^{ikx}\) with \(k\) is a fixed parameter.
Our goal is to find an approximation to \(\tilde{f}_n\).

Taking \(S=n(\zeta - q \eta)\), we can compute
\begin{gather*}
	\nabla (e^{iS} f_n(\rho, \eta)) = e^{iS} \left[i\nabla S f_n + \partial_q f_n \nabla q + \partial_\eta f_n \nabla \eta\right] \\
	\nabla S = n(\nabla \zeta - q \nabla \eta - \eta \nabla q)
\end{gather*}

Note that in the ballooning representation, we also have
\begin{gather*}
	\tilde{S} = n(\zeta - q(\eta - \theta_k)) \\
	\nabla(e^{i\tilde{S}} \tilde{f}_n(\theta_k, \eta)) = e^{i\tilde{S}} (i\nabla \tilde{S} \tilde{f}_n + \partial_\eta \tilde{f}_n \nabla \eta) \\
	\nabla \tilde{S} = n(\nabla \zeta - q \nabla \eta - (\eta - \theta_k) \nabla q)
\end{gather*}

Note furthermore we have the gyroaverage operator taking the eikonal approximation
\begin{gather*}
	\ev{e^{i\tilde{S}}\tilde{f}_n(\theta_k, \eta)} \approx e^{i\tilde{S}} J_0(k_\perp \rho) \tilde{f}_n(\theta_k, \eta) \\
	\vb{k}_\perp = \nabla \tilde{S}
\end{gather*}

\subsection{Up-down Symmetry}
One final thing we do is to modify \(\theta_0\) to account for up-down symmetry.
We compute \(v_r = \vb{v}_D \cdot \vu{r}\) and \(v_y = \vb{v}_D \cdot \vu{y}\) where \(\vu{y} = \vu{r} \times \vu{b}\) is the binormal vector.
We take \(\theta_0 = \operatorname{argmax}_{\theta} v_y\).

\subsection{Approximation by Gauss-Hermite Functions}

Finally, we approximate \(f_n(\rho,\eta)\) (equivalently \(\tilde{f}_n(\theta_k, \eta)\)) with complex-amplitude Gauss-Hermite functions.

Note that naively, if we try to extract \(\phi_n\) from the data \(\phi\), aliasing issues cause pollution in the toroidal mode number spectrum.
Instead, we first upsample the toroidal resolution using field-aligned interpolation (rather than numerically following field lines, we use constant \(\alpha=\zeta - q \theta\)), perform the FFT, then truncate in frequency space.
Additionally, we apply a Lanczos filter in the parallel direction to reduce high-frequency parallel noise -- not sure what its source is.

To compute gradients of \(\phi\), note the coordinate transform

Define \(v_D^q,v_D^\alpha\) by
\begin{equation*}
	\vb{v}_D\cdot \nabla = v_D^q \partial_q + v_D^\alpha \partial_\alpha = v_D^R \partial_R + v_D^Z \partial_Z + v_D^\varphi \partial_\varphi
\end{equation*}
Note that \(v_D^\varphi = \vb{v}_D \cdot \vu*{\varphi} / R\).

We have the following coordinate transforms
\begin{equation*}
	q = q(\psi) \qquad \alpha = \zeta - q \theta \qquad \eta = \theta
\end{equation*}
\begin{equation*}
	\begin{bmatrix}
		\dd{q} \\ \dd{\alpha} \\ \dd{\eta}
	\end{bmatrix}
	=
	\begin{bmatrix}
		q' & 0 & 0 \\
		-q' \theta & 1 & -q \\
		0 & 0 & 1
	\end{bmatrix}
	\begin{bmatrix}
		\dd{\psi} \\ \dd{\zeta} \\ \dd{\theta}
	\end{bmatrix}
	=
	\begin{bmatrix}
		q' & 0 & 0 \\
		-q' \theta & 1 & -q \\
		0 & 0 & 1
	\end{bmatrix}
	\begin{bmatrix}
		\psi_R & 0 & \psi_Z \\
		0 & 1 & 0 \\
		\theta_R & 0 & \theta_Z
	\end{bmatrix}
	\begin{bmatrix}
		\dd{R} \\ \dd{\varphi} \\ \dd{Z}
	\end{bmatrix}
\end{equation*}

\section{Miscellania}

\subsection{Derivative Operators}

Let's look at the parallel Laplacian
\begin{equation*}
	\nabla_\perp^2 
\end{equation*}

\end{document}
